---
title: "Assignment 6"
author: "Iden Watanabe"
date: "March 13, 2018"
output:
  html_document: default
  pdf_document: default
---

The purpose of this exercise is to load information from three different file types: HTML, XML, and JSON, and to see if the results are identical.  I picked mythology books for this assignment and handmade the HTML table, XML, and JSON files, which are uploaded to my GitHub.

```{r setup, message = FALSE}
library(XML)
library(knitr)
library(jsonlite)

html.url <- "https://raw.githubusercontent.com/EyeDen/data607/master/assignment6/books.html"
xml.url <- "https://raw.githubusercontent.com/EyeDen/data607/master/assignment6/books.xml"
json.url <- "https://raw.githubusercontent.com/EyeDen/data607/master/assignment6/books.json"

download.file(html.url, "books.html")
download.file(xml.url, "books.xml")
download.file(json.url, "books.json")
```

```{r html}
books.html <- as.data.frame(readHTMLTable("books.html"))
kable(books.html)
```

Aside from a minor issue with the column names, this is okay.  If we were cleaning this up, we'd have to split apart the authors for _The Odyssey_ and _Fairy Tales_.

Though a function exists as part of the `XML` library to read an XML format and convert it to a data frame, it is a bit too simplistic as it can't handle the multiple author tags for two of the books.  We'll have to try something else.  A [StackOverflow]["https://stackoverflow.com/questions/2067098/how-to-transform-xml-data-into-a-data-frame"] question gave me a solution, requiring the `plyr` library.

```{r xml}
books.xml <- ldply(xmlToList("books.xml"), data.frame)
```

This natively manages to split apart the authors, which `readHTMLTable` couldn't manage for HTML.  We can clean it up a bit, and remove columns we won't be needing.

```{r xml}
books.xml <- books.xml[, c(2, 3, 8, 4, 5, 6)]
kable(books.xml)
```

```{r json}
books.json <- as.data.frame(fromJSON("books.json"))
kable(books.json)
```

Reading in the JSON file, we get an output exactly like the HTML table, although the column names are a little neater.  Rather than separating the authors into new columns automatically, as the XML version had done, this leaves multiple entries in the same column.

So, no, the methods are not identical.